Staged Abstract Interpreters

### 2019-08-15 paper/30_stageconc.tex
revision

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ + (α → v)
        ev(e)(ρ_*)(σ_*)
    }; emit_compiled_clo(f)
  }
\end{lstlisting}

We have created a current function @f@ that takes a next-stage value
and store, and returns staged value. Then we use @emit_complied_clo@
to delay @f@ to a next Scala function, represented by the type
@Rep[Value]@. Inside of @f@, we

To make this happen, when
closing a lambda term, we need to specialize the body expression by
\textit{collpasing} the @AnsM@ monads to normal values. The
following function @close@ illustrates the idea: @f@ is a current stage function
but takes two next-stage values @v@ and $\sigma$, inside of which, we eagerly
collapse the monads to values by @ev(e)(ρ_*)(σ_*)@, which has type
@Rep[(Value, Store)]@. Then we generate the next-stage value of the compiled
closure using @emit_compiled_clo@, containing the compiled closure.

When applying a function in @ap_clo@, we also generate the next-stage value
for closure application through @emit_ap_clo@, assuming that the @rator@ is already an
IR of compiled closure, @rand@ is an arbitrary stage value, and $\sigma$ is the
latest store. @emit_ap_clo@ represents the result value of the application, which has type
@Rep[(Value, Store)]@. Note that @emit_ap_clo@ is an ordinary value from the
future, and there are no monads in the future, so we need to reify the value and
store back into the monad stack through @put_store@ and @yield@.

\begin{lstlisting}
  def emit_ap_clo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]): Rep[(Value, Store)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ <- get_store
    val res: Rep[(Value, Store)] = emit_ap_clo(rator, rand, σ)
    _ <- put_store(res._2)
  } yield res._1
\end{lstlisting}

@@

The method @emit_compiled_clo@ and @emit_ap_clo@ mentioned above produces
intermediate representations for the denoted operations in LMS. These IRs are
wrappers of the underlying staged data. We define these IR nodes using
@case class@ extending from @Def[T]@, meaning that they are next-staged definitions of type @T@.

\begin{lstlisting}
  type ValSt = (Value, Store)
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

After constructing the IR graph, LMS provides an @emitNode@ to generate code for
each kind of IR node. To generate code for @IRApClo@, we know that @rator@ is a
compiled closure @IRCompiledClo@, which contains a next-stage function @f@, so
we just need to generate code that invokes @f@ with @rand@ and @σ@ as arguments
provided. Similarly, we directly generate a next-stage value @CompiledClo(f)@
for @IRCompiledClo@.

\begin{lstlisting}
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
    case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar$rand, $\textdollar$σ)")
    case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar$f, $\textdollar$λ, $\textdollar$ρ)")
    ...
  }
\end{lstlisting}

>>>>

\begin{lstlisting}
  type ValSt = (Value, Store)
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[ValSt], λ: Lam, ρ: Exp[Env]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[ValSt] = { case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ + (α → v))
    }; emit_compiled_clo(f, λ, ρ)
  }
\end{lstlisting}

We have created a current-stage function @f@ that takes a next-stage value and
store, and returns staged value. Then we use @emit_complied_clo@ to delay @f@ to
a next-stage Scala function, represented by the type @Rep[Value]@. Inside of
@f@, we can access the evaluator via @ev@, but @ev@ produces a monadic value of
type @AnsM@, which can only exist at the current stage. To connect the
current-stage monadic values and future-stage grounded values, the evaluator
@ev@ is invoked on the body expression @e@ to be specialized w.r.t. @e@, and
then we \textit{collpase} the @AnsM@ monad to grounded values by providing the
prepared environment and store.

For function applications @ap_clo@, we have two \textit{next-stage} values @fun@
and @arg@, but what we can do on these values of @Rep@ type? In fact, @fun@ does
not have an intensional application operation we can use directly; but we know
that it is a staged value produced by the @close@ function, which represents a
next-stage Scala function. Therefore we can just generate a next-stage function
application, also in Scala. Since @ap_clo@ returns a monadic value of type
@Ans@, we can still use the monadic operation @get_store@ to obtain the latest
store object. Then we use @emit_ap_clo@ to generate a current-stage
representation of the future-stage application result, which is a pair of values
and stores @ValSt@. After getting that result @vs@, we use @put_store@ to
replace the store and reify the result to the current-stage monadic value.

\begin{lstlisting}
  def emit_ap_clo(fun: Rep[Value], arg: Rep[Value], σ: Rep[Store]): Rep[ValSt]
  def ap_clo(ev: EvalFun)(fun: Rep[Value], arg: Rep[Value]): Ans = for {
    σ  <- get_store
    vs <- lift[ValSt](emit_ap_clo(fun, arg, σ))
    _  <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

Compared with the unstaged version, we can observe a shift of using the
evaluation function @ev@ from application-time to value-representation-time: in
the unstaged interpreter, the invoke of @ev@ happens in @ap_clo@; while in the
staged interpreter, we eagerly apply @ev@ when denotating the lambda term to our
value domain.

@@

In the previous section, we merely show the type of @emit_compiled_clo@ and
@emit_ap_clo@ without a concrete implemtation. In this section, we brifely
discuss the IR node generated by them and sketch code generation.
We define the IR nodes @IRCompiledClo@ and @IRApClo@ using @case class@
extending from @Def[T]@. @Def[T]@ is a built-in type in the LMS framework,
representing next-stage definitions of type @T@.

\begin{lstlisting}
  case class IRCompiledClo(f: Rep[(ValSt) => ValSt], λ: Lam, ρ: Rep[Env]) extends Def[Value]
  case class IRApClo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]) extends Def[ValSt]
\end{lstlisting}

The IR nodes are manipulated by the LMS passes and finally generated to
next-stage Scala code. To generate code, LMS provides an @emitNode@ method for
programmers to control what is generated for each kind of node. Here we have
matched @IRCompiledClo@ and @IRApClo@ and will generate definitions for them.
For compiled closures, we put the function @f@ and its companying syntactic term
and environment to its next-stage value representation @CompiledClo@. For
applications, we have the knowledge that @fun@ is a @CompiledClo@ at the stage,
and it has a callable field @f@, so we generate a next-stage function
application @f(arg, σ)@. The @quote@s are invocations to generate code for the
other next-stage values.

\begin{lstlisting}
override def emitNode(sym: Sym[Any], rhs: Def[Any]) = rhs match {
  case IRCompiledClo(f, λ, ρ) => emitValDef(sym, s"CompiledClo($\textdollar${quote(f)}, $\textdollar${quote(λ)}, $\textdollar${quote(ρ)})")
  case IRApClo(rator, rand, σ) => emitValDef(sym, s"$\textdollar$rator.f($\textdollar${quote(rand)}, $\textdollar${quote(σ)})")
  ...
}
\end{lstlisting}

### 2019-08-15 paper/40_absinterp.tex
paper
revision

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time abstraction, but for the purpose of clearly
illustrating the idea, we will just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components, namely
abstract values, abstract environments, and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines the type
class of stage polymorphic lattices. Similar to @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to
annotate the binding-times, where @R[_]@ wraps @S@ and @Boolean@ in these
operations.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
  \begin{lstlisting}[style=small]
trait SPLattice[S, R[_]] {
  val ⊤: R[S];  val ⊥: R[S]
  def ⊑(l1: R[S], l2: R[S]): R[Boolean]
  def ⊔(l1: R[S], l2: R[S]): R[S]
  def ⊓(l1: R[S], l2: R[S]): R[S]
}
trait Lattice[S] extends SPLattice[S, NoRep]
  \end{lstlisting}
  \caption{trait \texttt{SPLattice} and \texttt{Lattice}} \label{fig:splattice}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.6\textwidth}
\begin{lstlisting}[style=small]
def SetLattice[T]: Lattice[Set[T]] = new Lattice[Set[T]] {
  lazy val ⊥: Set[T] = Set[T]()
  lazy val ⊤: Set[T] = error("No representation for ⊤")
  def ⊑(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf  l2
  def ⊔(l1: Set[T], l2: Set[T]): Set[T]  = l1 union     l2
  def ⊓(l1: Set[T], l2: Set[T]): Set[T]  = l1 intersect l2
}
\end{lstlisting}
  \caption{The power set instance for lattice} \label{fig:powerset}
\end{subfigure}
\end{figure}

In this section, we simply let @R[T] = T@ (i.e., @NoRep@) to instantiate the
unstaged lattice type class. A simple instance of lattice we used is the power
set abstract domain (shown in Figure \ref{fig:powerset}), where the ordering is subset relation, the join is union
operation, and the meet is the intersect operation.
Other lattices used in the remainder of the paper, such as products and maps, can be
implemented similarly or lifted from existing lattices element-wise or
point-wise. Naturally, abstract domains such intervals also can be implemented
in a stage polymorphic way.

@@

@@

% \cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12}

We follow the \textit{abstracting definitional interpreters} (ADIs) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter. We first widen the concrete values to @AbsValue@: numbers
are lifted to a singleton object @IntTop@ representing the set of all numbers;
closures remain the same. Then the type @Value@ is sets of @AbsValue@ (we abuse
the type names a little bit, as the generic interface defined the abstract type
@Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After settling the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, meaning an address may point to an over-approximated set of values
occurred at runtime.

@@

To prevent the abstract interpreter from falling into non-termination, the ADI
approach uses a two-cache mechanism. Here we first provide the type
definitions, and later we will describe the algorithm in detail. A @Cache@ is a
mapping from configurations @Config@ to sets of value-store pairs, where the
configuration is a triple of current expression being evaluated, environment and
store.

@@

As of now, the store is a mapping from addresses to sets of abstract values.
Therefore, when applying a function, for instance, @f(a)@, we may retrieve a set
of closures by dereferencing the address of @f@. We need to nondeterministically
apply all of the target closures.
Another nondeterminism happens when evaluating conditionals, we have to evaluate
both branches if the condition can not be determined. In other words, we will
have multiple computing results.

@@

This observation leads us to the type definition of @Result@: in a concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a set of such pairs
@Set[(Value, Store)]@ to represent multiple results. The results are also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

\paragraph{Monads for Abstract Interpretation} Now we incorporate the
nondeterminism and caching into the monad stack. For concrete evaluation, we
already have a reader effect for environments and a state effect for stores. We
further add three more effects based on it: the nondeterminism effect is
represented by the @SetT[M[_], A]@ monad transformer, inside of which, we have
a @ReaderT@ for a @in@ cache and a @StateT@ for an @out@ cache. \todo{highlight
where changed}

\begin{lstlisting}
  trait AbstractSemantics extends AbstractComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[SetT[ReaderT[StateT[IdM, Cache, ?], Cache, ?], ?], Store, ?], Env, T]
    ...
  }
\end{lstlisting}

The sketch implementation of the @SetT@ transformer is shown below. The value @run@
encapsulated by the monad is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and
@A@ is the element type of the list. The @flatMap@ method first unwraps the inner
monadic value @M[SetT[A]]@ and obtain a value of @Set[A]@, then applies @f@ to
each element in that list, and finally folds the results back to a single
@SetT[M, B]@.

\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] = ...
  }
\end{lstlisting}

\paragraph{Primitive Operations} Some of the primitive operations are changed
according to the new monad stack. A notable change is that the store update
operation joins the value with the existing values, for the same address. Under
abstract interpretation, we may have to explore the results from both branches
of a conditional expression for soundness. In @br0@, we combine the results
using @mplus@ from @MonadPlus@, which requires the values are join-semilattices
(in our case @Set@s and @Map@s).

@@

The value representation of lambda terms is still @CloV@, but we lift it to a
singleton set.

@@

For function applications, @ap_clo@ looks roughly the same as the concrete
interpreter. The difference is now the function @rator@ is a set of closures, so
we can not directly destruct it by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a set and lifts the elements into the
monad stack. Then we can still implement @ap_clo@ in monadic style, where the
@CloV@ comes from the @SetT@ and the nondeterminism is naturally handled.

\begin{lstlisting}
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    CloV(Lam(x, e), ρ) <- lift_nd[AbsValue](rator)
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been widely used in abstract
interpretation or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
Rosendahl:AbsIntPL}. Since the @in@ and @out@ cache are wrapped in the monad
stack, we first defined several methods to obtain or update the @in@ and @out@
cache through the monad stack.

@@

Figure \ref{fig:coind_cache} shows the co-inductive caching algorithm that
closes the open-recursive @eval@.
The idea is to first set up two caches called @in@ and @out@ both of type
@Cache@. Then similar to Kleene's iteration algorithm, we compute the analysis
result iteratively. The @in@ cache contains what we already have computed from
the last iteration, and the @out@ cache is what we would have computed after this
iteration joined with the contents from the @in@ cache.
When starting a new iteration, @in@ is set to be the result of the last
iteration, i.e., @out@; and @out@ is assigned to be empty.

During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation, if
yes, then the result is directly returned through the monads.

Otherwise, we first retrieve the results from @in@ (@⊥@ if not contained in
@in@), and put it into the @out@ cache.
Then we invoke @ev@ to compute the result for this iteration.
After the evaluation, the store @σ@ may have been changed. So we update the
@out@ cache with the new result, i.e., @(v, σ)@. Similar to store update, cache
update is also point-wise join.
Note that for the recursive call of @ev@, we also instrument it by putting
@fix(ev)@ as the first argument to @ev@.

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which means there is no more analysis result can be gained,
thus the iteration should end and we have reached the least fixed point.

\begin{figure}[t!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for {
            (v, σ) <- lift_nd[(Value, Store)](out(cfg))
             _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e)
            σ <- get_store
            _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\end{figure}

@@

Finally, we override the top-level @run@ method by running the monad with
the empty abstract environment $\rho_0$, initial store $\sigma_0$, and initial
caches @cache@$_0$.

>>>>

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We first present a
lattice framework with binding-time abstraction, as well as simple abstract
domains such as pwoer set lattice.
Then we show the abstract components, namely abstract values, abstract
environments, and abstract stores. The abstract interpreter we construct in this
section is similar to \citet{DBLP:journals/pacmpl/DaraisLNH17}'s. For
simplicity, it is also context/path/flow-insensitive by our choice. In Section
\ref{cfa}, we will instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

Abstract interpreters run on abstract domains, such abstract domains usually can
be presented by a lattice.
A complete lattice over set $S$ is a tuple of $\langle S, \sqsubseteq, \top,
\bot, \sqcup, \sqcap \rangle$, where $\top$ is the top element, $\bot$ is the
bottom element, $\sqsubseteq : S \to S \to Bool$ is the
ordering relation, $\sqcup: S \to S \to S$ is the join (least upper bound)
operator, and $\sqcap: S \to S \to S$ is the meet (greatest lower bound)
operator. The trait @SPLattice@ (Figure \ref{fig:splattice}) defines a type
class of stage polymorphic lattices. Similar to @MonadOps@ in Section
\ref{generic_if}, we introduce an additional higher-kinded type @R[_]@ to
annotate the binding-times, where @R[_]@ wraps datatype @S@ and @Boolean@ in
these operations.

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
  \begin{lstlisting}[style=small]
trait SPLattice[S, R[_]] {
  val ⊤: R[S];  val ⊥: R[S]
  def ⊑(l1: R[S], l2: R[S]): R[Boolean]
  def ⊔(l1: R[S], l2: R[S]): R[S]
  def ⊓(l1: R[S], l2: R[S]): R[S]
}
trait Lattice[S] extends SPLattice[S, NoRep]
  \end{lstlisting}
  \caption{trait \texttt{SPLattice} and \texttt{Lattice}} \label{fig:splattice}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.6\textwidth}
\begin{lstlisting}[style=small]
def SetLattice[T]: Lattice[Set[T]] = new Lattice[Set[T]] {
  lazy val ⊤: Set[T] = error("No representation for ⊤")
  lazy val ⊥: Set[T] = Set[T]()
  def ⊑(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf  l2
  def ⊔(l1: Set[T], l2: Set[T]): Set[T]  = l1 union     l2
  def ⊓(l1: Set[T], l2: Set[T]): Set[T]  = l1 intersect l2
}
\end{lstlisting}
  \caption{The power set instance for lattice} \label{fig:powerset}
\end{subfigure}
\end{figure}

In this section, we simply let the binding-time type @R[T] = T@ (i.e., @NoRep@)
to instantiate an unstaged lattice type class. An example of lattice we used in
the rest of the paper is the power set abstract domain (shown in Figure
\ref{fig:powerset}), where the ordering is subset relation, the join is union
operation, and the meet is the intersect operation. The bottom element of a
power set is empty set, and we don't have a top representation for power set.
Other lattices used in the the paper, such as products and maps, can be
implemented similarly or lifted from existing lattices element-wise or
point-wise. Non-relational abstract domains such intervals also can be
implemented in a stage polymorphic way.

@@

We follow the \textit{abstracting definitional interpreters} (ADI) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter.
% \cite{DBLP:conf/icfp/HornM10, DBLP:journals/jfp/HornM12}

@@

We first widen the concrete values to @AbsValue@: numbers are lifted to a
singleton object @IntTop@ representing the set of all numbers; closures remain
the same. Then the type @Value@ is a set of @AbsValue@ (we abuse the type names
a bit, as the generic interface already defined the abstract type @Value@).
The address space is constrained to be finite to ensure the computability of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After defining the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, indicating that an address may point to an over-approximated set of
values occurred at runtime.

@@

To prevent the abstract interpreter from falling into non-termination when
computing the fixed-point, the ADI approach uses a co-inductive mechanism
consisting of two caches that remember the input and output of functions.
Here we first provide the necessary type definitions, and later we will describe
the algorithm in detail. A @Cache@ is a mapping from configurations @Config@ to
sets of value-store pairs, where the configuration is a triple of current
expression being evaluated, environment and store. Intuitively, a cache memoizes
the corresponding values and stores for a given program configuration.

@@

Now, the store is a mapping from addresses to sets of abstract values. This can
be justied by the approximation nature of abstract interpretation and
nondeterminism in the abstract interpreter.
First, when analyzing a conditional experssion, we may not have enough
information on which branch will be taken, thus a sound treatment is to
nondeterministically explore both branches, i.e., compute the join of two path.
Second, when applying a function, for instance, @f(a)@, the possible lambda
terms of @f@ is computed approximately. As a result, we may retrieve a set of
closures when dereferencing the address of @f@ and need to apply all of them.

\paragraph{Monads for Abstract Interpretation} Compared with the concrete
interpreter that uses reader and state effects, the abstract interpreter
further introduces a nondeterminism effect and another reader and state effect
for the two caches. The nondeterminism effect is represented by the
@Set[M[_], A]@ monad transformer, where @M@ is the inner monad being
transformed, and @A@ is the type of elements in the set. We use a @ReaderT@ for
one cache that will not be changed during one fixed-point iteration, and use a
@StateT@ for another cache that will be constantly updated during analysis.
\citet{DBLP:journals/pacmpl/DaraisLNH17} discuss different permuatation of the
moand stack for abstract interpretation. The following @AnsM@ type shows the
monad stack (we use \hl{light gray} to highlight what have been changed from
the monad stack of concrete interpretation):

\begin{lstlisting}[escapechar=!]
trait AbstractSemantics extends AbstractComponents {
  type R[T] = T
  type AnsM[T] = ReaderT[StateT[!\hl{SetT}![!\hl{ReaderT}![!\hl{StateT}![IdM, !\hl{Cache}!, ?], !\hl{Cache}!, ?], ?], Store, ?], Env, T]
  ...
}
\end{lstlisting}

Similar to the concrete scenario, we sketch the implementation of @flatMap@ for
@SetT@ transformer and omit the rest. The value @run@ encapsulated by the monad
is of type @M[SetT[A]]@, where @M[_]@ is the inner monad, and @A@ is the element
type of the set. The @flatMap@ method takes a function that transforms each
element of type @A@ from @M[Set[A]]@ into a @SetT[M, B]@ value, and then folds
all the transformed values into a single monadic value of type @Set[M, B]@.

\begin{lstlisting}
  case class SetT[M[_]: Monad, A](run: M[Set[A]]) {
    def flatMap[B](f: A => SetT[M, B]): SetT[M, B] =
      SetT(Monad[M].flatMap(run) { (s: Set[A]) =>
        s.foldLeft(SetT.empty[M,B])((acc, a) => acc ++ f(a)).run
      }); ... }
\end{lstlisting}

Such monad stack scheme leads to a different type of result type. In the concrete
setting, we only have a reader monad (for environment) and a state monad (for
store), and the reader effect is not persistent to the final result. Together
with the value produced by the interpreter, the final result type is a pair of
@Value@ and @Store@. In the abstract setting, we have a @SetT@ inside of the
environment monad and store monad, which becomes the container type of the pair
of values and stores, i.e., @Set[(Value, Store)]@. Also note that the reader and
state monad for caches live internally of the nondeterminism monad, therefore
the cache type of state monad becomes adjacent to the set type.

@@

\paragraph{Primitive Operations} The primitive operations are changed to adapt
the new monad stack scheme and value domains. A notable change is that the store
update operation joins the value with the existing values, for the same address.
As what mentioned before, we may have to explore both branches for conditionals:
in @br0@ method, we combine the results using @mplus@ from @MonadPlus@, which
requires the value domains are join-semilattices (in our case @Set@s and
@Map@s).

@@

The value representation of lambda terms is still a defunctionalized closure
@CloV@, but we lift it to a singleton set domain.

@@

For function applications, @ap_clo@ looks the same as the concrete interpreter.
The difference is now the first argument @rator@ is a set of closures (remember
that @Value@ is now an alias of @Set[AbsValue]@), so we can not directly
destruct it by pattern matching. Instead, we use an auxiliary function @lift_nd@
that takes a set and lifts the elements into the monad stack. Then we can still
implement @ap_clo@ in monadic style, where the closures come from the @SetT@ and
thus the nondeterminism can be naturally handled. The light gray code shows what
is added from the concrete @ap_clo@.

\begin{lstlisting}[escapechar=!]
  def lift_nd[T](vs: Set[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    !\hl{CloV(Lam(x, e), $\rho$) <- lift\_nd[AbsValue](rator)}!
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- local_env(ev(e))(ρ + (x → α))
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been used in other abstract
interpretater or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
  Rosendahl:AbsIntPL}. The idea is to use a @in@ cache and a @out@ cache during
the depth-first evaluation, where the @in@ cache stores the result from last
iteration, and @out@ cache is used for updates in current iteration. In the next
iteration, the last @out@ cache will be served as @in@ cache, and an empty cache
is plugged-in to the @out@ slot. Once the @out@ cache does not contain anything
new than the @in@ cache, the fixed-point is reached.

In out monad stack, the @in@ and @out@ cache modeled by a reader effect and
state effect, respectively. We first defined several methods to manipulate
the two caches through the monad stack; the implementations are elided.

@@

The co-inductive caching algorithm is implemented as a instrumentation over
@eval@ (Figure \ref{fig:coind_cache}), and it also closes the open recursion.
Initially, the two caches are both empty.
During the iteration, we first check whether @out@ contains the configuration
@cfg@, representing the current desired computation. If it does, then the result
is directly returned through the monads.
Otherwise, we first retrieve the results from @in@ (@⊥@ if not contained in
@in@), and join it into the @out@ cache.
Then we invoke @ev@ to compute the result for this iteration, where @ev@ takes
@fix(ev)@ as a self reference.
After the evaluation, the store @σ@ may have been changed. We update the
@out@ cache with the new result, i.e., a pair of value and store @(v, σ)@.
Similar to store update, cache update is also a point-wise join.

The iteration terminates when the resulting @out@ cache is equivalent to the
input @in@ cache, which indicates that there is no more fact can be gained, thus
the iteration should end and we have reached the least fixed point.

\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for { // ask if out already contains the desired result
            (v, σ) <- lift_nd[(Value, Store)](out(cfg))
             _ <- put_store(σ)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e)
            σ <- get_store
            _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\end{figure}

@@

Finally, we override the top-level @run@ method by running the monadic value
with the empty abstract environment $\rho_0$, initial abstract store $\sigma_0$,
and initial caches @cache@$_0$.
